---
title: "Cyclist-Bike-Share-Analysis"
author: "Sourabh Marne"
date: '2022-03-10'
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ASK PHASE

Q.1 What is the problem you are trying to solve?

Ans: Our main aim is to find marketing strategies to convert casual riders into annual members.

Q.2 How can your insights drive business decisions?

Ans: Our insights will help the marketing team increase the annual members.

## PREPARE PHASE

1.  The data is located in a kaggle dataset which is stored in csv files month by month.
2.  The dataset contains entire population and there is no bias
3.  The dataset ROCCCs as it is reliable, original, comprehensive, current and cited.

#### Loading the libraries

```{r}
library(tidyverse)
library(lubridate)
```

1.  Tidyverse is used for data transformation.
2.  Lubridate is used for working with dates and time efficiently.

## PROCESS PHASE

Here we will prepare data for data analysis.

##### Reading the bikes data from all 12 months

```{r}
df1 <- read.csv("C:/Users/hp/Desktop/Google Data Anlytics Capstone/CSV files/202004-divvy-tripdata.csv")
df2 <- read.csv("C:/Users/hp/Desktop/Google Data Anlytics Capstone/CSV files/202005-divvy-tripdata.csv")
df3 <- read.csv("C:/Users/hp/Desktop/Google Data Anlytics Capstone/CSV files/202006-divvy-tripdata.csv")
df4 <- read.csv("C:/Users/hp/Desktop/Google Data Anlytics Capstone/CSV files/202007-divvy-tripdata.csv")
df5 <- read.csv("C:/Users/hp/Desktop/Google Data Anlytics Capstone/CSV files/202008-divvy-tripdata.csv")
df6 <- read.csv("C:/Users/hp/Desktop/Google Data Anlytics Capstone/CSV files/202009-divvy-tripdata.csv")
df7 <- read.csv("C:/Users/hp/Desktop/Google Data Anlytics Capstone/CSV files/202010-divvy-tripdata.csv")
df8 <- read.csv("C:/Users/hp/Desktop/Google Data Anlytics Capstone/CSV files/202011-divvy-tripdata.csv")
df9 <- read.csv("C:/Users/hp/Desktop/Google Data Anlytics Capstone/CSV files/202012-divvy-tripdata.csv")
df10 <- read.csv("C:/Users/hp/Desktop/Google Data Anlytics Capstone/CSV files/202101-divvy-tripdata.csv")
df11 <- read.csv("C:/Users/hp/Desktop/Google Data Anlytics Capstone/CSV files/202102-divvy-tripdata.csv")
df12 <- read.csv("C:/Users/hp/Desktop/Google Data Anlytics Capstone/CSV files/202103-divvy-tripdata.csv")
```

##### Binding all the dataframes together

```{r}
rides <- rbind(df1,df2,df3,df4,df5,df6,df7,df8,df9,df10,df11,df12)
head(rides)
```

##### Removing Duplicates

```{r}
rides <- rides[!duplicated(rides$ride_id), ]
print(paste("Removed", nrow(rides) - nrow(rides), "duplicated rows"))
```

##### Converting the "started_at" and "ended_at" columns from characters to Timestamps

```{r}
rides$started_at <- lubridate::ymd_hms(rides$started_at)
rides$ended_at <- lubridate::ymd_hms(rides$ended_at)
```

##### Converting ride time in minutes

```{r}
rides <- rides %>% mutate(ride_mins = as.numeric(rides$ended_at - rides$started_at) / 60)
summary(rides$ride_mins)
```

##### Separating year and month

```{r}
rides <- rides %>% mutate(year_month = paste(strftime(rides$started_at, "%Y"),
              "-",strftime(rides$started_at, "%m"),
                 paste("(",strftime(rides$started_at, "%b"), ")", sep="")))
unique(rides$year_month)
```

##### Extracting the week days

```{r}
rides <- rides %>% mutate(weekday = paste(strftime(rides$ended_at, "%u"), "-", strftime(rides$ended_at, "%a")))
unique(rides$weekday)
```

##### Extracting the ride hour

```{r}
rides$hour <- lubridate::hour(rides$ended_at)
unique(rides$hour)
```

## ANALYZE PHASE

```{r}
summary(rides)
```

##### Finding number of casuals vs number of annual members

```{r}
rides %>% group_by(member_casual) %>% summarise(count = length(ride_id),
            'Percentage' = (length(ride_id) / nrow(rides)) * 100)
```

##### Plotting number of casuals vs annuals

```{r}
ggplot(rides, aes(member_casual, fill=member_casual)) + geom_bar() +
  labs(x="Casuals vs Annual Members", title="Casuals vs Members Distribution")
```

##### Monthly Distribution of the Data

```{r}
rides %>%
  group_by(year_month) %>%
  summarise(count = length(ride_id),
            'Percentage' = (length(ride_id) / nrow(rides)) * 100,
            'members_p' = (sum(member_casual == "member") / length(ride_id)) * 100,
            'casual_p' = (sum(member_casual == "casual") / length(ride_id)) * 100,
            '% Difference' = members_p - casual_p)
```

```{r}
rides %>%
  ggplot(aes(year_month, fill=member_casual)) +
  geom_bar() +
  labs(x="Month", title="Chart 02 - Distribution by month") +
  coord_flip()
```

##### Insights:

1.  There are more riders in 2020.
2.  August has the biggest number of data points(18% of the dataset).
3.  We have more members than riders in every month.

##### Let us observe how the temperature affects number of riders

```{r}
chicago_mean_temp <- c(-3.2, -1.2, 4.4, 10.5, 16.6, 22.2, 24.8, 23.9, 19.9, 12.9, 5.8, -0.3)
month <- c("001 - Jan","002 - Feb","003 - Mar","004 - Apr","005 - May","006 - Jun","007 - Jul",
           "008 - Aug","009 - Sep","010 - Oct","011 - Nov","012 - Dec")
```

```{r}
data.frame(month, chicago_mean_temp) %>%
  ggplot(aes(x=month, y=chicago_mean_temp)) +
  labs(x="Month", y="Mean temperature", title="Mean temperature for Chicago (1991-2020)") + geom_col()
```

1.  As we can see the temperature highly affects the number of riders in every month
2.  The colder the weather, the lesser the bike riders

##### Week Day Distribution

```{r}
rides %>%
  group_by(weekday) %>% 
  summarise(count = length(ride_id),
            '%' = (length(ride_id) / nrow(rides)) * 100,
            'members_p' = (sum(member_casual == "member") / length(ride_id)) * 100,
            'casual_p' = (sum(member_casual == "casual") / length(ride_id)) * 100,
            '% Difference' = members_p - casual_p)
```

```{r}
ggplot(rides, aes(weekday, fill=member_casual)) +
  geom_bar() +
  labs(x="Weekday", title="Distribution by weekday") +
  coord_flip()
```

##### Insights:

1.  The highest volume of riders is on weekends.
2.  Saturday being the highest.
3.  More casual riders are seen on weekends.

##### Hourly Distribution

```{r}
rides %>%
  group_by(hour) %>% 
  summarise(count = length(ride_id),
            '%' = (length(ride_id) / nrow(rides)) * 100,
            'members_p' = (sum(member_casual == "member") / length(ride_id)) * 100,
            'casual_p' = (sum(member_casual == "casual") / length(ride_id)) * 100,
            '% Difference' = members_p - casual_p)
```

```{r}
rides %>%
  ggplot(aes(hour, fill=member_casual)) +
  labs(x="Hour of the day", title="Hourly Distribution of the day") +geom_bar()
```

##### Insights:

1.  There are more bikers in the afternoon.
2.  There are more members in the morning from 5 am to 11 am.
3.  There are more casuals in the afternoon from 11 pm to 4 am.

##### Weekends vs Mid Week Distribution

```{r}
rides %>%
  mutate(type_of_weekday = ifelse(weekday == '6 - Sat' | weekday == '7 - Sun', 'weekend', 'midweek')) %>%
  ggplot(aes(hour, fill=member_casual)) +
  labs(x="Hour of the day", title="Distribution by hour of the day in the midweek") +
  geom_bar() +
  facet_wrap(~ type_of_weekday)
```

##### Insights:

1.  Weekends have mre casuals from 11 am to 6 pm.
2.  There is huge rise in weekdays from 5 pm to 6 pm, then it drops.
3.  On weekends there is smooth flow which rises in the afternoon.

##### Type of Rideable bikes

```{r}
rides %>%
  group_by(rideable_type) %>% 
  summarise(count = length(ride_id),
            '%' = (length(ride_id) / nrow(rides)) * 100,
            'members_p' = (sum(member_casual == "member") / length(ride_id)) * 100,
            'casual_p' = (sum(member_casual == "casual") / length(ride_id)) * 100,
            '% Difference' = members_p - casual_p)
```

```{r}
ggplot(rides, aes(rideable_type, fill=member_casual)) +
  labs(x="Rideable type", title="Distribution of types of bikes") + geom_bar() +   coord_flip()
```

##### Insights:

1.  Docked bikes have highest volume of riders.
2.  Members prefer the classic and electric bikes.

## SHARE PHASE

### Key Insights:

1.  Members have the biggest proportion of the dataset than casuals.
2.  There are more riders at the last semester of 2020. August recorded the biggest count of data points which was 18% of the dataset.
3.  Temperature heavily influences the number of rides in each month.
4.  Weekends have highest number of riders.There are more riders in the afternoon.

### Members vs Casual Riders:

1.  Members have highest volume of data, except for saturdays.
2.  Weekends have more casuals than members.
3.  There are more members in the morning from 5am to 11 am.
4.  There are more casuals from 11pm to 4am.
5.  There is an increase from 6am to 8am on weekdays for members. Next big rise is from 5pm to 6pm.
6.  Members prefer classic bikes.
7.  Casuals have more riding time than members.

### Conclusion:

1.  Bikes are used as an recreational activity on weekends.
2.  Temperature affects the number of riders.
3.  Members use bikes for daily activities like going to work.

## ACT PHASE

#### Now the marketing teams can use these insights for increasing the number of members and converting casuals to annual members.
